---
title: "Class 2"
output: html_notebook
---

## Load the pulse data
```{r echo = F}
pulse_data <- read.csv('Pulse experiment response.csv')
str(pulse_data)
```

```{r}
head(pulse_data)
```

## Print the data
```{r}
print('Sitting')
subset(pulse_data,Posture == 'Sitting')$Pulse
print('Standing')
subset(pulse_data,Posture == 'Standing')$Pulse
```












## Plot data
```{r,fig.width=4,fig.height=5}
library(ggplot2)
set.seed(1)
ggplot(pulse_data,aes(x=Posture,y=Pulse)) + geom_jitter(size=2,width = .1,alpha = 0.5)
```

## Try a histogram of each group
Each bar gives the number of observations in each bin (range of the data). All bins have the same size.
We can convert the bins from counts to percentages by adding `aes(y=..density..)` to `geom_histogram()`. 
```{r}
library(ggplot2)
breaks = seq(30,130,by=5)
ggplot(pulse_data,aes(x=Pulse)) + 
  geom_histogram(binwidth = 5, center = 2.5,color = 'black',fill = 'grey70') + 
  facet_grid(Posture~.)

# same, but re-scaled to percentages
ggplot(pulse_data,aes(x=Pulse)) + 
    geom_histogram(aes(y=..density..),binwidth = 5, center = 2.5,color = 'black',fill = 'grey70') + 
    facet_grid(Posture~.) + 
    ggtitle('Probability plot')
```

## Try a boxplot of each group
- The heavy bar in the center is the **sample median**
- The box brackets the **Interquartile range**, ie, from the point greater than 25% of the data to the 
point greater than 75% of the data.
- Vertical lines then extend out to the smallest and biggest data points, as long as they are within 1.5x the 
interquartile range from the box.
- Any points outside this range are plotted as outliers
```{r,fig.width=4,fig.height=5}
# compare to jitter_plot
ggplot(pulse_data,aes(x=Posture,y=Pulse)) + geom_jitter(size=2,width = .1,alpha = 0.5)  +
     stat_summary(fun.y='mean', colour="darkred", geom="errorbar", aes(ymax = ..y.., ymin = ..y..))

# boxplot
ggplot(pulse_data,aes(x=Posture,y=Pulse)) + geom_boxplot()
```

## Run a t-test
```{r}
standing_obs = subset(pulse_data,Posture == 'Standing')$Pulse
sitting_obs = subset(pulse_data,Posture == 'Sitting')$Pulse
t_test = t.test(standing_obs, sitting_obs, paired = F, var.equal = T)
t_test
```


# View data.frame
```{r}
head(pulse_data)
```












## sample estimates:
mean of x mean of y
```{r}
mean(standing_obs)
mean(sitting_obs)
```


## qnorm
```{r}
pnorm(3) - pnorm(-3)
qnorm(.975)
```


## Confidence interval:
```{r}
ybar_1 = mean(standing_obs)
ybar_2 = mean(sitting_obs)
n1 = sum(pulse_data$Posture == 'Standing')
n2 = sum(pulse_data$Posture == 'Sitting')
s2_1 = var(standing_obs)
s2_2 = var(sitting_obs)

# pooled estimate of population variance (assuming both pops have same variance)
# pooling counts each individual equally
s2 = ((n1-1)*s2_1 + (n2-1)*s2_2)/(n1+n2-2)

# SED estimate using pooled s2
sed = sqrt(s2/n1 + s2/n2)

# t_crit
tc = qt(0.05/2,n1+n2-2,lower.tail=F)

# confidence interval
ybar_1-ybar_2 + c(-1,1)*sed*tc
```

