---
title: "HW 1"
author: "YOUR NAME"
date: "DUE Thursday January 16"
output: 
 html_notebook:
    toc: true
    toc_float: true
    number_sections: true
---

Use this R Notebook document to answer the questions and document your work. 
Enter the R code used to answer each question in the corresponding R code chunk. 
Write any textual explanations **outside** of the chunks. 
Attempt to clean up your code as much as possible so that only the necessary lines remain. 

When you are done:

1. Select 'Run All' from the 'Run' dropdown menue.
2. Save (File -> Save)
3. Click 'Preview' to bring up the `HW1.nb.html` file. Check through this to make sure it rendered correctly.
4. Upload the files: `HW1.nb.html` and `HW1.Rmd` to Canvas.

To make your assignments easy to auto-grade, please follow these requirements:

- Do not change the directory in your code (ie `setwd()`). Place the file in a folder with any associated data files
provided and compile the file in that directory. This allows me to re-run your code if needed.
- **Hard-code** file-names into your code - don't use `file.choose()` to select the file.
- When a variable is provided to hold your answer for a specific problem, be sure to assign your answer to that variable. 
Ex. See question 1.1 below. You can directly type in a number, or assign the value using R commands.

---

# Part 1

## 10 numbers were generated by the computer from a normal population with mean=18 and standard deviation=5. 
[4 points]

| 15 | 18 | 24 | 28 | 14 | 19 | 25 | 17 | 20 | 18 |
|----|----|----|----|----|----|----|----|----|----|

Using `R` as a calculator (i.e., use only the commands you'd find on a hand calculator: `+ - * / sqrt`), calculate:

- mean
```{r}
# use this code block to enter your answer. Be sure to run the code block so that your answer prints out below
```

- standard deviation
```{r}

```

- coefficient of variation (CV)
```{r}

```

- standard error of the mean (SEM)
```{r}

```


## Add 20 to each value and indicate the effect on the mean, standard deviation, coefficient of variation, and SE. Comment on the differences from above. You may now use the `mean` and `sd` functions.
[4 points]
```{r}
# mean

# SD

# CV

# SEM

```

> ENTER YOUR RESPONSE HERE

## These questions are worded poorly. This should be re-written to ask about differences in means, rather than differences in observations.

## Now, we have generated three additional samples of 10 numbers from the same distribution (mean = 18, sd=5). 
[4 points]
The data are in the R object called `data_1`. Calculate the mean, SD, CV and SE for each of the four samples. 

```{r echo=FALSE}
# DO NOT enter your own code here - it won't be printed in the output.

# This code enters the data into a data.frame called `data_2.3`. 
data_1 <- data.frame(
              Sample_1 = c( 15 , 18 , 24 , 28 , 14 , 19 , 25 , 17 , 20 , 18 ),
              Sample_2 = c( 28 ,  7 , 13 , 18 , 16 ,  8 , 16 , 17 , 16 , 18 ),
              Sample_3 = c( 16 ,  6 , 21 , 19 , 16 , 21 , 18 , 16 , 19 , 13 ),
              Sample_4 = c( 20 , 27 , 12 , 28 , 26 , 18 , 12 , 21 , 18 ,  9 )
            )

# You can print a data.frame to the screen / Notebook by typing it's name, and then CTRL+ENTER
data_1

# A data.frame is like a spreadsheet. You can access each column by referring to its name:
data_1$Sample_1

# or by it's index:
data_1[,1]
```

For this question, you can use R functions like `mean` or `sd`. 
Note: There are no built-in functions for CV or SE
```{r}
# mean
```
```{r}
# SD
```
```{r}
# CV
```
```{r}
# SE

```

## Calculate and compare:
[4 points]

- the mean of the four sample means. Compare this value to the overall mean (17.775).
Note: to calculate the mean using the function `mean`, enclose the values in a vector: `mean(c(1,3,4,2))`

```{r}

```
- the standard deviation of the four sample means. Compare this value to the average of the four standard deviations from above.
Note: to calculate the sd using the function `sd`, enclose the values in a vector: `sd(c(1,3,4,2))`
```{r}

```

- the standard deviation of the four sample means. Compare this to the theoretical standard error (standard deviation of sample means with sample size=10). 
```{r}

```
    
Explain why the standard deviation of the four means and the theoretical standard error of the mean (n=10) are smaller than the average of the four standard deviations.

> ENTER YOUR RESPONSE HERE

## The following table was derived by subtracting Sample 1 from Sample 2 (S2-S1) and subtracting Sample 3 from Sample 4 (S4-S3). 
[4 points]

Calculate the mean and variance for each derived sample, and then find the average of these statistics.  
Compare these values with the theoretical values expected from the subtraction of two random variables, 
based on the true population parameters defined above.

```{r echo=FALSE}
data_2 <- data.frame(
              S2_minus_S1 = data_1$Sample_2 - data_1$Sample_1,
              S4_minus_S3 = data_1$Sample_4 - data_1$Sample_3
            )
data_2
```


```{r}
# Mean of means

```
```{r}
# mean of variances

```

> ENTER YOUR RESPONSE HERE

---

# Part 2 

[Hint: With questions like these, it is very helpful to draw a figure first]

## Assume Y has a Normal distribution with mean=20 and variance = 10. Calculate the following:
[5 points]

Note: R's normal distribution functions use SD, not variance!

- Of 100 individuals sampled from this population, approximately how many would you expect to be smaller than 25?
```{r}

```

- Of 100 individuals sampled from this population, approximately how many should be within 1.5 times the population's standard deviation? 
```{r}

```

- Find the expected 75th quantile of the distribution of Y (the value greater than 75% of the population)
```{r}

```

- With a sample size of 15 from this population, what is the probability that the sample mean will be less than 19?
```{r}

```

- With a sample size of 15, what is the probability that the sample mean will be within 1.5 times the population's standard deviation? 
```{r}

```

---

# Part 3
[10 points]

Describe an experiment that you are currently running, one that you have run in the past, or will run soon, 
or make up an experiment that is relevant to your research. Use the terms we discussed in the first class. 
Cover the following items. A couple sentences or a brief list is fine. The goal of this is to get you thinking
about the issues in experimental design, and to help me understand the range of questions students are studying.
    
## Research question: 
What is the overal goal of the experiment?

> ENTER YOUR RESPONSE HERE

## Treatment structure: 
What treatments are you imposing? How are you imposing them? Are they qualitative or quantitative?
Are you using controls?

> ENTER YOUR RESPONSE HERE

## Design structure: 
What are your experimental units? What is the reference population (the broader set of all eu's you want to generalize about)?
How do you assign treatments to eu's? How many eu's per treatment?
How do you randomize? What confounding factors are you concerned about? What sources of variation do you expect will affect your results?

> ENTER YOUR RESPONSE HERE

## Response structure: 
What traits / quantities will you measure? How many measurements will you take per eu? 
How will you measure? What precautions will you take to prevent biases?

> ENTER YOUR RESPONSE HERE

## Are you OK wih me using / adapting this as an example in class / labs / exams?

> ENTER YOUR RESPONSE HERE

---
