---
title: "Quick Intro to R"
author: Mitchell Feldmann and Daniel Runcie
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---
# Quick Intro to R
For the purposes of PLS 205 and beyond, learning how to manipulate and analyze data in R will be a very valuable skill. 
This document designed to introduce you to some of the basic ideas that are widely applicable. 
We start with bringing data into our environment & loading packages before moving into simple calculations, vectors, data.frames, lists, data management, and finally plotting.

## R scripts
R scripts are text files (like this one) that contain all R commands necessary to run an analysis.
Scripts work just like entering commands sequentially into the **Console**, but provide a history
of what you did.

### The '#' symbol:
```{r}
# The pound sign/hashtag is very important in R, even though it isn't a function. 
# The '#' denotes that the text following it is a comment. 
# Anything written after a '#' will not be treated as an instruction for R to run. 
# In RStudio, you can see that text after a '#' changes color to indicate it's a comment, and not code. 
# You can use comments to explain in plain English what your code is doing, and why. 
# Your code should have lots of comments.  
# Comments can come after a command on the same line or on their own line. Nothing is returned from this chunk.
```

### Code blocks
Below is a code block. In a code block, all text is interpreted as `R` code that should be evaluated.
Any line that is not an R command should be started with a `#`

```{r}
'Hello'
3+5
# A comment
```


### install.packages() and library():
There are thousands of free packages available that extend R's functionality. 
Each package includes one or more functions that do specific tasks.
Packages can be installed with the function:
`install.packages()`, or using the **Packages** window in RStudio.

The following code will install most of the packages we will use in this course
```{r}
# install.packages(c('lme4','lmerTest','car','multcomp','lsmeans','pbkrtest','ggplot2')) # remove the `#` symbol to run once. Then add it back so it doesn't run again!
```

Once packages are installed, you need to load them each time you want:
```{r}
library(lme4)
library(ggplot2)
```


## Use R as a mathematical and boolean calculator:
Run each of the commands below separately by putting your cursor on the line and clicking **Run** above
```{r}
# as you would expect the standard computer calculator symbols apply: 
# + is addition
# - is subtraction
# * is multiplication
# / is division
# ^ or ** are exponents

2*3
(1+2+3)/3
2^3
```

R has many functions for doing math:
```{r}
sqrt(4) # square root
pi
exp(1) # the natural exponential
log(exp(1)) # natural log
log2(2) # log base 2
log10(10) # log base 10
```

R can evaluate logical statements:
```{r}
5 > 3    # > is greater than, < is less than
6 == 5   # two equal signs is the logical test (is equal to). (One equal sign is assignment. see below)
5 != 3   # not equal to
5 >= 5   # greater than or equal to

# compare text
"bookshelves" == "a totally different character"
"apples" != "oranges"
"apples" != "Apples" # R is always case-sensitive
```

---

## Working with data

All data that we will work with in this class will be spreadsheet-like,
meaning that it will be arranged in columns, with rows as observations, and columns as descriptors or numbers.
A simple data set is provided called `iris_data.csv`. Open this data in Excel and look at it.
Note that each plant is give one row, and is labeled with a species and four measurements

### loading data
The main functions for reading data into `R` are: `read.csv`, `read.tsv`, or `read.delim`, depending on
how your data are encoded.

```{r}
iris <- read.csv('iris_data.csv')  # default function
iris <- read.csv('iris_data.csv',header = TRUE)  # explicitly specify that the first row of the data contains column names
iris <- read.delim('iris_data.csv',header = TRUE,sep=',')  # explicitly specify that entries are separated by commas
```

> The above commands have three important parts:
>
> 1. The **function** (read.csv or read.delim). Functions do things. Here, it reads data.
> 2. The function call is given **arguments** that tell it what to do. Here, we provide a filename, and other paramters to control it's behavior
> 3. The **assignment operator** `<-`. This tells R to take the output of the function (here, the loaded data), and assign it
> to the **object** called `iris`. Once the data is assigned to the object `iris`, it is held in memory and we can access it with other functions.
> You can see all objects you have currently in memory in RStudio's **Environment** window

RStudio also provides a neat tool for importing data. 
Choose `File -> Import Dataset -> From Text (base)`, select the file name, then use the dialog
box to describe the file. This will generate an R command for you and run it in the console.
Simply copy that command, and paste it into your document here:

```{r}
# Copy the import command that RStudio generated for you and paste it here in your script

```

### Viewing data
Now that the data are loaded and assigned to the object `iris`, we can view it in several ways:

```{r}
iris
# alternative: print(iris)
```

Simplying typing the name of the object and running the code **prints** out the whole object to the screen.
(You can close this view by clocking the little `x` icon in the top-right corner of the output)


```{r}
str(iris)
```
`str` is a **function** that provides a concise summary of the data. 
Note that `iris` is a data.frame, with 150 observations and 5 variables.
The first four variables are all numbers, and the 5th is a Factor. We'll discuss the differences later

```{r}
head(iris)
tail(iris)
```
`head` and `tail` are **functions** that show you the first or last 6 entries of the data

RStudio also provides the function `View()`. It doesn't work inside an R Notebook, but if you type:
`View(iris)` in the console, it will bring up something that look like a spreadsheet in a new window.

```{r}
summary(iris)
```
`summary` is similar to `str` but calculates numerical summaries of each column.

### Accessing data
Now that your data are loaded, we can access it in several ways:

```{r}
iris[1,]
```
This gives the first row of the data

```{r}
iris$Sepal.Length
# alternatives: iris[,1], iris[,'Sepal.Length'], iris[['Sepal.Length]]
```
This accessess all the values in the `Sepal.Length` column

Commands can be combined:
```{r}
summary(iris$Sepal.Width)
```
```{r}
head(iris$Petal.Length)
```

Data can be accessed and assigned to a new **object**
```{r}
setosa <- subset(iris,Species == 'setosa')
str(setosa)
```

> Notes:
>
> 1. The function `subset` takes two arguments: a data.frame (iris), 
> and then a logical vector specifying which **rows** of the data.frame to select
> 2. The output is a new data.frame with only the observations of the **setosa** species listed

We can now reference the setosa data specifically:
```{r}
mean(setosa$Sepal.Length)
```

### Plotting data
R has many functions for plotting data. 

There are two main flavors: **base graphics**, and **ggplot** graphics. base graphics are quicker,
but ggplot graphics look much better and are faster for some tasks

#### scatterplot
```{r}
# base graphics
plot(x = iris$Sepal.Length,y = iris$Sepal.Width,            # Note: commands can extend across lines
     xlab = 'Sepal Length (mm)', ylab = 'Sepal Width (mm)',
     main = 'Iris Sepals')
# ggplot
library(ggplot2)
ggplot(iris,aes(x=Sepal.Length,y=Sepal.Width)) + 
  geom_point(aes(color = Species)) +
  ggtitle('Iris Sepals')
```

#### Boxplots
```{r}
# base
boxplot(iris$Petal.Length~iris$Species,ylab = 'Petal.Length',ylim = c(0,8),main = 'Iris petals')
# ggplot
ggplot(iris,aes(x=Species,y=Petal.Length)) + geom_boxplot() + ylim(c(0,8))

```

#### Histograms
```{r}
# base
hist(iris$Sepal.Width,breaks = 10)
# ggplot
ggplot(iris,aes(x=Sepal.Width)) + geom_histogram(bins = 10,alpha = 0.7,color='white')
```


---

## R programing
Running analyses in R is more than just loading data and writing the right model. 

A good analysis will perform and document all steps in your workflow, from data cleaning, pre-processing,
modeling, interpretation, and display.

These tasks require learning a bit about how to write scripts or programs.

Below, we demonstrate some key tools for working in the `R` language:


### Getting help
1. Search in the **Help** pane.
2. Type `?` or alternatively `help` and `??` (or alternatively `help.search`). Use `?` (or `help(yourcommand)`) if you know the exact command name and use `??yourcommand` (or `help.search("yourcommand")`)  if you don't know the command name, R will return a list of commands matching (somehow) your search.
```{r}
?read.csv	# the same as help(read.csv)
??lmer	# the same as help.search("lmer")
```
3. Search online. Pretty much any question you might have has been asked by someone in some forum. Learning how to find these answers is a very useful skill.


### Calling R functions
Functions are the workhorses of R. They can do simple things, like find the mean of some numbers, load data, or evaluate complicated models.
Pretty much everything you do in R involves using (or writing your own) functions, so it's important to understand how they work.

Here are some functions:
```{r}
mean(c(1,2,3)) # calculates the mean
sort(c(10,3,2,5)) # sorts numbers
qnorm(p = 0.95,mean = 10,sd = 3) # finds the 0.95 critical value of a normal distribution with mean = 10 and sd = 3
```

Functions take *arguments* as input, and then *return* values as output. If you call a function by itself like above, the result is 
added directly to your R Notebook in the chunk output. However, you can *assign* this output to a variable with the `=` symbol
```{r}
m3 = mean(c(1,2,3))
m3
```

Most functions require that you provide them data to work on. You can see what a function requires to run in it's help page by typing `?mean`, 
or highlighting the word and typing `F1`. The help page lists the arguments of a function in RStudio's help pane. 
Each argument is named. Some arguments have default values. You don't have to pass an argument that has a defualt value if you want to use that value. 

Compare how we called `mean` and `qnorm` above. Note that we specified the argument names in `qnorm`, but not for `mean`. 
Argument names aren't strictly required as long as you pass the values in the correct order. 
But it's always a good idea to use the name so that your code is easier to read, and it's easier to find mistakes.

The qnorm function could also have been called like this:
```{r}
qnorm(p = 0.95,m = 10,s = 3) 
```
*Partial matching* of argument names works to save you time (as long as it's unamibiguous which argument you mean, R will guess). But this is a bit dangerous too!


### tab auto-complete:
```{r, eval=FALSE}
# the tab key can be used to auto-complete words in R. This can save a lot of typing!

# It works for the columns of a data.frame (the standard object for holding data files)
# try hitting tab with the cursor after `iris$Sep` below:
iris$Sepal.

# It works for variable names you've used already in this document:
# try below;  
seto

# It works for any R function in any package you've already loaded:
read.cs

# It also works with files and directories:
'iris_

```

### Vectors
Vectors are like data-frames that are one-dimensional (either a column or a row of a data.frame)

Vectors are useful for grouping similar objects together

Vectors can be created with the `c()` contatenation function:

```{r}
# vector of characters
furniture <- c("sofa", "chair", "table", "desk")
furniture

# vector of numbers
heights <- c(85.4, 98, 45, 55, 3.7)
heights
```

Elements of vectors can be named
```{r}
furniture_sizes <- c(sofa = 83, chair = 30, table = 102)
furniture_sizes
```

Elements of vectors can be accessed with brackets `[]`
```{r}
furniture[2]
heights[3:5]
furniture_sizes['sofa']
```

Vectors can also be extracted from data.frames:
```{r}
observation_1 <- iris[1,]
observation_1

data_1 <- iris$Sepal.Length
data_1
```

`observation_1` and `data_1` are both vectors

> We'll use vectors like this to store answers in homeworks and exams.

When data is stored in a vector, we can apply the same operation to each element of the vector
```{r}
heights
heights + 10
heights > 80
```


### Loops
We won't use them much in this class, but loops are what make programing powerful.
The idea of a loop is to repeat the same command over and over again.
The following command draws 10 random numbers 100 times, saving the mean of the 10 each time in a vector
```{r}
means <- rep(0,100) # prepare a vector for the results
for(i in 1:100){  # this for loop repeates the commands below between the brackets for each number between 1 and 100
  # each time through, the object `i` gets assigned to the current iteration number
  sample <- rnorm(n = 10, mean = 20, sd = 3)  # we use `rnorm` to draw 10 values from a normal distribution. See ?rnorm
  means[i] <- mean(sample)  # we calculate the mean of the 10 sampled numbers, and assign the value to the ith element of the vector `means`
}
```

Lets make a histogram of these means:
```{r}
hist(means)
```


