---
title: "Lab 1"
output: 
 html_notebook:
    toc: true
    toc_float: true
---

In this lab, we'll replicate the sampling demonstration [at this website](http://www.zoology.ubc.ca/~whitlock/kingfisher/KFhomepage.htm).
This will demonstrate using R to draw random numbers, calculate means, standard deviations, standard errors, and confidence intervals, 
and perform t-tests and power calculations.

You can go through this lab using the `Lab1.nb.html` document by highlighting the code in the code blocks and copying it into an R terminal.
Or, you can go through the R Notebook file directly, running code in the code blocks directly.

---

## Part 1: Sampling from a normally distributed population

The first part of the demonstration explores the variability of sampling individuals from a larger population 
and measuring them to learn about the population's mean and variance.

This is based on [this website](http://www.zoology.ubc.ca/~whitlock/kingfisher/SamplingNormal.htm).

### Normal distribution calculations in R

In the demonstration, we are interested in the length of fish in a pond (the population). 
We assume that the fish lengths follow a normal distribution with mean: $\mu = 106$mm, and standard deviation: $\sigma = 30$mm.

R has four functions for the normal distribution. Look at the help page by typing `?dnorm`. They are:

- `rnorm(n, mean = 0, sd = 1)` This generates `n` independent random draws from the distribution.
- `qnorm(p, mean = 0, sd = 1, lower.tail = TRUE)` This is the *quantile function*. 
It replaces the critical value table. 
The default (with `lower.tail=TRUE`) is to give you the 100x`p`% quantile of the distribution, 
i.e., the critical value greater that 100x`p`% of random draws a normal distribution with specified mean and standard deviation. 
To get the upper quantile, change `lower.tail=FALSE`.
- `pnorm(q, mean = 0, sd = 1, lower.tail = TRUE)` 
This is the *distribution function*. 
This is the probability of drawing a value less (greater if `lower.tail=FALSE`) than `q` from a normal distribution with the specified mean and standard deviation.
- `dnorm(x, mean = 0, sd = 1)` 
This calculates the *density* (height) of the normal distribution with specified mean and standard deviation at the value `x`.

Pretty much every distribution in R has the same four functions, 
specified with `rDIST`, `qDIST`, `pDIST` and `dDIST`, with `DIST` replaced by the distribution name. 
But some, like the *t*, *chi square*, or *F* distribution have additional *arguments*, such as the degrees of freedom of the distribution.

### Step 1: Sample 1 individual

We can ask `R` to simulate the sampling of a single individual from this population using the `rnorm` function:
```{r}
rnorm(n = 1, mean = 106, sd = 30)
```

> What size did you get? Did you get the same number as the person sitting next to you? 

> Try running the code several times. Do you get different numbers?

**Note** We usually specify normal distributions based on their variance ($\sigma^2$) rather than their standard deviation ($\sigma$), 
but `R` always specifies normal distributions based on the standard deviation

### Step 2: Sample 10 individuals

You could run the above code 10 times and write down the fish sizes from each sample. But R has two ways of doing this automatically:

Option 1: Use a loop:
```{r}
sample_of_fish = rep(NA, times=10)  # prepare a vector to record the 10 fish
for(fish_number in 1:10){   # a for-loop runs the code inside the {brackets} several times, each time changing the variable 'fish_number'
  print(fish_number)  # prints the variable 'fish_number' for this run of the code
  sample_of_fish[fish_number] = rnorm(n = 1, mean = 106, sd = 30)  # samples a fish, and records it in the vector
}
sample_of_fish
```

Option 2: Just ask for 10 fish!
```{r}
sample_of_fish = rnorm(n = 10, mean = 106, sd = 30)
sample_of_fish
```


Calculate the mean of your sample
```{r}
mean(sample_of_fish)
```

> How close is this to the *true* mean of the population? Compare your mean to those around you.

Make a histogram of these fish lengths
```{r}
hist(sample_of_fish, xlim = c(0,300),prob = T)

# add a line for the sample mean. 
abline(v = mean(sample_of_fish),lwd = 2)

# add a red line for the true population mean
abline(v = 106, col=2,lwd=2)

# add a line showing the true population distribution
x = seq(0,300)
lines(x,dnorm(x = x,mean = 106, sd = 30))
```

**Note**: For the above code to work, you have to run the whole code chunk at once (use the green arrow at the top-right of the chunk)


### Step 3: Mean from many samples

To explore this process, pretend you could repeate this sampling process many times. 

Let's ask R to draw 100 samples of size 10 from this population

For each sample, we'll calculate it's mean

Again, use a for-loop
```{r}
sample_means = rep(NA,100)
for(i in 1:100){ # do 100 experiments
  sample_of_fish = rnorm(n = 10, mean = 106, sd = 30)  # sample 10 fish
  sample_means[i] = mean(sample_of_fish) # record the sample mean
}
sample_means
```

Calculate the mean of the 100 samples
```{r}
mean(sample_means)
```

Calculate the standard deviation of the 100 samples
```{r}
sd(sample_means)
```


The expected distribution of the sample means is a normal distribution with $\mu = 106$ and $\sigma = 30 / \sqrt{n} = `9.49`$


**Note**: In R Notebooks, *hover* the mouse over an equation to see it rendered!

Make a histogram of the sample means, and compare to the expected distribution
```{r}
hist(sample_means,xlim = c(0,300),prob = T)

# add line for expected distribution
x_positions = seq(0,300)  # create a list of x-values
y_positions = dnorm(x,mean = 106,sd = 30/sqrt(10))  # at each x-value, calculate the distribution at that point
lines(x_positions,y_positions)
```

### Questions:

> If you drew more samples of size 10, how would the above histogram change? Try it!

> If you drew larger samples (say 30), how would the above histogram change? Try it!

> If the fish were actually less variable (say sd = 10), how would the above histogram change? Try it!

---

## Part 2: Confidence intervals for the mean

This second part of the demonstration explores the creation and interpretation of confidence intervals.

This is based on [this website](http://www.zoology.ubc.ca/~whitlock/kingfisher/CIMean.htm).

Confidence intervals are used to express our uncertainty about an estimate (ex. population mean). 
The interval gives a range of plausible values for the true mean. 

Calculating confidence intervals requires using R's function for the **t-distribution**

### Student's t-distribution

The t distribution functions work the same as the normal distribution functions, except it has two additional parameters:

- `df` The degrees of freedom of the t distribution
- `ncp` A non-centrality parameter. Ignore this. We won't use it.

### Step 1: Sample 10 fish from the original population
```{r}
n = 10
mu = 106
sd = 30
sample_of_fish = rnorm(n=n, mean = mu, sd = sd)
```

### Step 2: Calculate the sample mean and sample sd
```{r}
sample_mean = mean(sample_of_fish)
sample_mean

sample_sd = sd(sample_of_fish)
sample_sd
```

The sample mean is $\bar{y} = `r sample_mean`$, and the sample sd is $s = `r sample_sd`$.

### Step 3: Calculate a confidence interval

The formula for a confidence interval is: $CI = \bar{y} \pm t_{1-\alpha/2,df}\mbox{SE}$.

$t_{1-\alpha/2,df}$ is the **critical value** of the *t* distribution at the $\alpha$ level, with $df$ degrees of freedom.

The Standard Error is $\mbox{SE} = \frac{s}{\sqrt{n}}$

Calculate the standard error, the t critical value for $\alpha = 0.05$, and then the confidence interval
```{r}
SE = sample_sd / sqrt(n)
SE

alpha = 0.05
t_crit = qt(p = 1-alpha/2,df = n-1)  # this gives the value greater than 1-0.05/2 = .975 possible values from the t-distribution
t_crit

CI = sample_mean + c(-t_crit * SE, t_crit*SE)
CI
```

> Does the confidence interval include the true mean? How many confidence intervals IN THE CLASS Do Not include the true mean? 
How many would you expect to not include the true mean?

Calculate the length of the confidence interval:
```{r}
length_CI = CI[2] - CI[1]
length_CI
```

### Questions:

> If you drew more samples of size 10, how would you expect the length to change? Try it!

> If you drew larger samples (say 30), how would you expect the length to change? Try it!

> If the fish were actually less variable (say sd = 10), how would you expect the length to change? Try it!
 

---

## Part 3: The t-test

Lastly, we'll explore the one-sample t-test, and compare it to the confidence interval

We can perform a one-sample t-test to test the null hypothesis $\mu = 106$. Read the help file for the function `?t.test`
```{r}
t_test_result = t.test(x=sample_of_fish,  # your data
       y = NULL,          # NULL, because it's a one-sample test (testing mean against known value)
       mu = mu,           # mean of the null hypothesis
       alternative = 'two.sided',     # two-sided alternative hypothesis
       conf.level = 1-alpha    # confidence level (desired frequence)
       )
t_test_result
```

> Do you reject the null hypothesis? Is this consistent with your confidence interval?

**Note:** the `t.test` function gives you a confidence interval. Does this match the one you calculated? You can directly extract those numbers like this:
```{r}
t_test_result$conf.int
```
