---
title: "HW5"
output: html_notebook
---

```{r}
library(ggplot2)
library(ggplot2)
library(emmeans)
```


# Question 1

Consider the same experiment that you studied in HW4:

> To study the effect of burial time on germination and dormancy of *Hordeum murinum ssp. leporinum* (Hare Barley), 
> overwintering seeds were placed in bags in a field, and then removed and tested at six time intervals (0, 30, 60, 90, 120, and 150 days after > burial). The design was a completely randomized design, with 5 replicate bags per burial time.
> Each bag contained 200 seeds. At the designed time, each bag was opened, and then the seeds were split evenly between two 
> germination trays, and the number of germinating seeds was measured in each tray.

Considering what we discussed in lecture this week, re-consider your choice over experimental units. 
Is "Bag" really an appropriate experimental unit for the effect of burial time?

## 1 Provide a critique of this experimental design
Start by re-making the design table, and carefully considering the experimental units.
Is it sufficient to answer the intended question?

**Design**: Completely randomized design

| Structure | Variable   | Type        | # levels | Experimental Unit |
|-----------|------------|-------------|----------|-------------------|
|           |            |             |          |                   |
|           |            |             |          |                   |
|           |            |             |          |                   |
|           |            |             |          |                   |

> ENTER YOUR RESPONSE HERE

## 2 What would you recommend to the researcher in terms of how she should report the results?
What type of analysis would you recommend?
How should the interpret the treatment effect they observe?

> ENTER YOUR RESPONSE HERE


# Question 2

A researcher conducted a study comparing the effect of different ground covers on nitrous oxide fluxes from apple orchard soils. She randomly assigned the inter-row spaces of an orchard block to one of four treatments: 1) no cover (Bare); 2) fescue (Grass); 3) white clover (Legume); and, 4) fescue-clover mix (Mix). Each treatment level was replicated in 6 plots. After rain events, emissions were collected from each plot, analyzed, and the emitted nitrous oxide was calculated in g N~2~O ha^-1^ d^-1^. Average emissions for each plot are reported below.

| Rep |  Bare | Grass | Legume |  Mix  |
|:---:|:-----:|:-----:|:------:|:-----:|
|  1  | 44.79 | 21.59 |  20.08 | 21.48 |
|  2  | 53.71 | 21.58 |  16.76 | 24.46 |
|  3  | 68.69 | 25.73 |  15.15 | 25.47 |
|  4  | 70.46 | 32.08 |  17.72 | 27.02 |
|  5  | 88.11 | 27.60 |  15.13 | 18.52 |
|  6  | 95.53 | 22.11 |  19.98 | 21.54 |

```{r echo=FALSE}
NO_data = read.csv('HW5_data.csv')
str(NO_data)
```

## 3 Describe the design of this experiment in detail.

**Design**: Completely randomized design

| Structure | Variable   | Type        | # levels | Experimental Unit |
|-----------|------------|-------------|----------|-------------------|
|           |            |             |          |                   |
|           |            |             |          |                   |
|           |            |             |          |                   |

## 4 Test the necessary assumptions of ANOVA. Report on the results.

```{r}
par(mfrow=c(1,2))
```

> ENTER YOUR RESPONSE HERE

## 5 Perform an appropriate transformation of the data. Re-check the assumptions of ANOVA, and report the results.

```{r}
par(mfrow=c(1,2))
```

> ENTER YOUR RESPONSE HERE

## 6 Provide estimates of effects of the various ground covers relative to Bare ground on both the original and transformed data ($\alpha = 0.05$). 
Describe and compare the results of both models. Be sure to provide units for the effect sizes on both scales.

```{r}
```

> ENTER YOUR RESPONSE HERE

## 7 Provide estimates and confidence intervals for the means of the 4 treatment.
**Are the means greater or less than the means estimated on the un-transformed scale?**

```{r}
```

> ENTER YOUR RESPONSE HERE