---
title: "HW3"
output: 
 html_notebook:
    toc: true
    toc_float: true
    number_sections: true
---
Use this R Notebook document to answer the questions and document your work. 
Enter the R code used to answer each question in the corresponding R code chunk. 
Write any textual explanations **outside** of the chunks. When we grade, we will *hide* the code and just look at the R output and your written answers.
Attempt to clean up your code as much as possible so that only the necessary lines remain. 

When you are done:

1. Select 'Run All' from the 'Run' dropdown menue.
2. Save (File -> Save)
3. Click 'Preview' to bring up the `HW1.nb.html` file. Check through this to make sure it rendered correctly.
4. Upload the `HW1.nb.html` to Canvas.

---

# Question 1

In lab, we analyzed an experiment with subsamples. Multiple soil samples were taken per plot.
It turns out that in a balanced experiment where evey experimental unit has the same number of subsamples,
you will get *exactly* the same answers if you first average the subsamples into a single mean response per plot,
and then analyze the averaged data.

In this question, I want you to repeat the analyses presented in Lab3 after first averaging the subsamples per plot.

The following code loads the Sample-level data from lab and averages the Samples per Plot:

```{r}
clover <- read.csv('Clover_data.csv')
clover_means <- aggregate(NLevel ~ Plot + Strain,clover,FUN = mean)
str(clover_means)
```

## Explain why the soil samples are subsamples, not experimental units

> Enter your response here.

## Provide a new design table for the averaged data

**Design**: Completely randomized design

| Structure | Variable | Type | # levels | Experimental Unit |
|-----------|----------|------|----------|-------------------|
| Treatment |          |      |          |                   |
| Design    |          |      |          |                   |
| Response  |          |      |          |                   |

## Write the model for the averaged data and fit it with the appropriate model function
```{r}

```

## Show appropriate diagnostic plots and assess whether the data satisfy the model assumptions
Compare the plots to those generated in lab.
Note: To replicate the `qqplot` and `Scale-Location` plot shown in Lab, you can use `plot(model,which=2:3)`
```{r}
par(mfrow=c(1,2))
```

> Enter your response here

## Provide 95% confidence intervals for the effect of each of the 5 new strains relative to the control strain
Compare your results to the results from the full data in R.
```{r}

```

> Enter your response here


---

# Question 2
A Pharmacy is testing 5 new cough syrups. Two are derived from a steroid compoind, and two from dexamethasone. 
The researchers include the current standard drug as a control, and label the syrups A-E so that the administering doctors do not
know the identity of the drug (a double-blind study). Each drug was administered to 5 different people. 
Below is the average coughs/hour of the replications:

| Drug            |       |       |       |       |       |
|-----------------|-------|-------|-------|-------|-------|
| A_Steroid       | 17.13 | 31.26 | 24.74 | 29.08 | 36.69 |
| B_Steroid       | 20.82 | 45.82 | 29.52 | 27.35 | 24.08 |
| C_Dexamethasone | 29.08 | 53    | 48.69 | 42.13 | 36.63 |
| D_Dexamethasone | 31.26 | 30.17 | 41.04 | 17.13 | 11.69 |
| E_Old_standard  | 55.17 | 49.74 | 48.62 | 37.78 | 35.61 |

```{r}
cough_data = read.csv('Coughing.csv')
str(cough_data)
```

## Describe the design of this experiment in detail.

**Design**: Completely randomized design

| Structure | Variable | Type        | # levels | Experimental Unit |
|-----------|----------|-------------|----------|-------------------|
| Treatment |          |             |          |                   |
| Design    |          |             |          |                   |
| Response  |          |             |          |                   |

## Visualize the data using a boxplot
Does it appear visually that there are differences among drugs? Do you see any data quality issues?
```{r}

```

> ENTER YOUR RESPONSE HERE

## Assess whether the assumptions of normality and equality of variances are satisfied by this data.
```{r}
```

> ENTER YOUR RESPONSE HERE

## Estimate the effect of each of the drugs relative to the old standard 
How strong is the evidence that any of the new drugs improve on the old standard?
Provide appropriate 95% confidence intervals for each estimate
Use `ref = X` in the call to emmeans to control which treatment level is treated as the control.
```{r}

```

> ENTER YOUR RESPONSE HERE

## The company wants to move forward with a single drug - can you recommend one of the drugs over all the others?
This asks you to declare one drug to perform better than all others. 
- If you choose to recommend a drug, be very specific about what criteria you used, and declare your confidence in this conclusion.
- If you choose not to recommend any specific drug, declare why you are not able to declare the best-performing drug the winner?
```{r}

```

> ENTER YOUR RESPONSE HERE

## Now, the company decides they do want to release a new Steroid drug, and a new Dexamethasone drug. 
The following code splits the data into two separate datasets for each class of drug.
Directly compare the two steroid drugs, and separately compare the two Dexamethasone drugs. 
**Can you recommend either steroid drug? Either Dexamthasone drug?**
Use alpha = 0.05
```{r}
steroid_data <- subset(cough_data,Type == 'Steroid')
Dexamethasone_data <- subset(cough_data,Type == 'Dexamethasone')

```
Your analysis should differ qualitatively from your analysis above (two drugs that were not distinguishable above are distinguishable here)

- Why would this analysis differ from the previous analysis? Differences may include: the set of questions, the calculation of SE, and the calculation of p-values / Confidence intervals
- Are the above p-values reasonable to report as evidence for a difference between either of the two pairs of drugs?

> ENTER YOUR RESPONSE HERE


